Reduced plasticity and opportunity for selection in terrestrial ectotherm populations under climate change

### 2023-12-04 ms.qmd
Writing intro

Variable thermal environments are expected to result in strong selection pressures that lead to adaptation or the evolution of phenotypic plasticity -- both of which are considered critical for population resilience to human-induced climate change [@Seebacher2015-wi]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable. Despite this theoretical expectation, empirical support for this prediction is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity). Reversible forms of phenotypic plasticity, such as acclimatization, may not be expected to adhere to such predictions.

@@

Applying new effect sizes that allow us to make use of powerful meta-analytic models we: 1) re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity; 2) test whether the opportunity for selection on physiological traits changes as temperatures rise, by applying new effect sizes that capture changes in physiological trait variance, and 3) test whether climate variability and predictability explain a populations acclimation capacity and affect changes in trait variance.

>>>>

Climate change is expected to result in warmer and more variable thermal environments for many organisms globally [@Suarez2023]. Greater thermal variability is expected to result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered critical for population resilience to human-induced climate change [@Seebacher2015-wi]. Nonetheless, the extent to which plasticity versus genetic adaptation will allow organisms to adapt to new environmental conditions is not yet clearly established. Without adaptive genetic or plastic responses high rates of extinction are predicted [@Cahill2012]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable. Despite this theoretical expectation, empirical support for this prediction is scant [], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [], and/or the prediction is likely only supported for specific plastic responses (i.e., active and developmental plasticity). Reversible forms of phenotypic plasticity, such as acclimatization, may not be expected to adhere to such predictions.

@@

Reversibly plastic responses, such as acclimitisation, are expected to provide organisms with resilience to climate change by allowing them to adjust their physiology to match the thermal environment they are exposed to [@Seebacher2015-wi].

Here, we apply new effect sizes to: 1) re-evaluate the degree to which aquatic and terrestrial ectotherms are capable of physiological plasticity; 2) test whether the opportunity for selection on physiological traits changes as temperatures rise and 3) test whether climate variability and predictability explain a populations acclimation capacity and affect changes in trait variance.
